WASP-4 is Accelerating Towards the Earth

### 2020-01-29 paper/ms.tex
update paper

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  The period change could be caused by tidal orbital decay, orbital
  precession, or light-travel time effects.  In this work we present
  new radial velocity measurements of WASP-4, acquired with
  Keck-HIRES.  The data show that the system is accelerating towards
  the Earth at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
  m}\,{\rm s}^{-1}\,{\rm yr}^{-1}$.  This implied period decrease
  probably explains WASP-4's changing orbital period as a light-travel
  time effect.  By combining the radial velocities with new speckle
  imaging limits, we find that system's acceleration is likely caused
  by a 10-200$\,M_{\rm Jup}$ companion with semi-major axis between
  10-100$\,$AU.  Based on statistics from
  \citet{knutson_friends_2014}, about 1 in 6 hot Jupiters are expected
  to show period decreases commensurate with WASP-4 due to
  acceleration by outer companions.  Continued long-term radial
  velocity monitoring of hot Jupiters is therefore essential to
  distinguish tidal orbital decay from line-of-sight accelerations.
\end{abstract}

@@

The issue of whether and how the orbits of hot Jupiters decay due to
tidal interaction with their host stars has drawn significant
attention over the past 25 years
\citep{rasio_tidal_1996,levrard_falling_2009,matsumura_tidal_2010}.
Numerous observations have also been performed (CITE).  These efforts
are now yielding fruit: by combining transit and occultation timing
with radial velocity measurements, \citet{yee_orbit_2020} have shown
that the data for the hot Jupiter WASP-12b are most compatible with
orbital decay.

The present study highlights a point that, though obvious, has perhaps
not yet received due attention.  The point is that observational
programs aimed at identifying orbital decay in hot Jupiters will be
crippled without long-term radial velocity monitoring programs.  The
reason is simple: when the timescale of orbital period change exceeds
the observing baseline, the first deviation from constant periodicity
is always quadratic in the transit number.

While orbital decay may be common (CITE).  line-of-sight accelerations
could be even {\it more} common (CITE Knutson 14, Bryan 16).  Both
manifest to first order identically in transit times (up to the sign
of the line-of-sight acceleration) -- as a non-zero period derivative.
Specifically, \citet{knutson_friends_2014} showed that X\% of hot
Jupiters show some form of radial acceleration in radial velocity
time-series with baselines of YY years.  Assuming a separation
distribution of (WHATEVER), this implies that ZZ\% of hot Jupiters
will {\it always} show a changing orbital period, over timescale of
say 10 years.

% The abundance of giant planets
% exterior to Super-Earth sized systems is also high (CITE Bryan 2019),
% but this is not so much of a concern to us, as their planetary masses
% and therefore tidal timescales are typically expected to be at least
% an order of magnitude shorter than hot Jupiters.

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that appears to be decreasing by about 10 milliseconds
per year.  We discovered the timing variations using data from TESS
and a decade of ground-based observations \citep[][hereafter
\citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.  Thereafter,
\citet{southworth_transit_2019} reported 22 new transit times for the
system, and found an updated decay rate of $\dot{P} = -9.2 \pm 1.1$
milliseconds per year. The \citeauthor{southworth_transit_2019} decay rate
was $\approx$3$\sigma$ less rapid than that found by
\citetalias{bouma_wasp4b_2019}, but the conclusions of the
studies were otherwise the same.  A separate study by
\citet{baluev_2019} reported additional archival transit light curves
of WASP-4b, most notably from the TRAPPIST archive, as well as from
select other observers.  \citet{baluev_2019} pointed out that
when using lower-precision subsets of the available transit data,
the case for a decreasing period worsened.  We agree with this
statement.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the five available
HIRES radial velocities suggested a weak ($<2\sigma$) linear trend
\citep{knutson_friends_2014}.  Our new measurements reveal a
line-of-sight acceleration of $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m\,s^{-1}\,yr^{-1}}$.  This
translates to an expected period decrease from the light-travel time
effect of -5.9 milliseconds per year---about commensurate with what
is observed from transits.  In the following, we avoid the term
``R{\o}mer delay'', which typically signifies arrival time
delay due to observatory motion. We are discussing the Doppler effect
seen by a stationary observer for a source with constant line-of-sight
acceleration.

In the following, \S~\ref{sec:observations} collects the available
transit data and presents the new radial velocity and speckle imaging
observations.  \S~\ref{sec:analysis} analyzes the data, and finds that
they yield a picture in which the WASP-4 system is accelerating
towards our line-of-sight, likely due to the pull of a brown or
M-dwarf companion.
\S~\ref{sec:discussion} places this result in the broader context of
orbital decay searches, and points out that line-of-sight
accelerations, {\it i.e.}, ``false positive orbital decay signals'',
are relatively common for the hot Jupiter population.
\S~\ref{sec:conclusions} offers brief concluding remarks.

@@

{\bf Table~N} lists the transit times we collected for our analysis.
We include data from the peer-reviewed literature for which (a) the
analysis was based on observations of a single transit, (b) the
midpoint was fitted as a free parameter, and (c) the time system
specified both the leap second correction (TDB or UTC) and also
whether any barycentric or heliocentric corrections had been
performed.

@@

Additional timimg measurements have also recently been reported by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations taken professional and amateur observers.
We included twelve of their ``high quality'' transit times acquired by
TRAPPIST (six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had or had not been performed.

@@

The complete set of radial velocity observations is given in {\bf
Table~M}.  Along with the 2010-2019 HIRES observations, there are also
earlier measurements from CORALIE and HARPS.  Following
\citetalias{bouma_wasp4b_2019}, we included the CORALIE measurements
from \citet{wilson_wasp-4b_2008} and \citet{triaud_spin-orbit_2010},
using the homogeneous radial velocities calculated by the latter
authors. We included the HARPS values reported by
\citet{pont_determining_2011} and \citet{husnoo_observational_2012}.
We omitted the HARPS data points taken over three nights by
\citet{triaud_spin-orbit_2010} for Rossiter-McLaughlin observations
because they have a systematic offset from the remaining datasets.

@@

We considered two models for the observed transit times.  The first
model assumes a constant orbital period on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes that the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the observed times.  This helps
to reduce the covariance between $t_0$ and $P$.  A third possible
model that we did not consider for reasons that will become apparent
is a precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fit constant period model.  The best-fitting constant-period
model has $\chi^2 = 276$, 91 degrees of freedom, and $\chi^2_{\rm
red}=3.0$.  The best-fitting quadratic model has $\chi^2 = 183$, 90
degrees of freedom, and $\chi^2_{\rm red}=2.0$.  The difference in the
Bayesian information criteria (BIC) between the linear and quadratic
and models is $\Delta {\rm BIC} = 89.1$, strongly favoring the
quadratic case \citep{kass_bayes_1995}.

@@

%FIXME: might just do the example calculation?
One approach to analyzing such data would be to introduce a fudge
factor that inflates the transit measurement uncertainties, and lowers
the reduced $\chi^2$.  We do not think that such an approach is
warranted, since it would not change the result that the quadratic
model is strongly preferred.

The best-fit period derivative in the quadratic model is
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$3$\sigma$ larger than the rate
of period decrease we found earlier \citep[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of
missing coverage filled in by \citeauthor{southworth_transit_2019}'s
and \citeauthor{baluev_2019}'s observations.  The other best-fit
transit timing model parameters are reported in {\bf Table~X}.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit
    Keplerian orbit of WASP-4b has been subtracted.  The linear trend
    inferred from the RV data is shown with a black line, with
    $1\sigma$ errors in gray.  The trend that would be needed to
    produce the period decrease seen in transits ($\dot{P}= - 8.64 \pm
    0.89~{\rm ms}\,{\rm yr}^{-1}$) is indicated with the purple dotted
    line.  The four new RV measurements presented in this work
    increased the significance of the radial velocity trend from
    $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period derivative expected from
radial velocities is
\begin{align}
  \dot{P}_{\rm RV}
  = \frac{\dot{v}_{\rm r} P}{c}
  = -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.

An important consideration is whether the measured RV trend is
correlated with stellar activity.  We investigate this by analyzing
WASP-4's emission in the Ca II H \& K lines, as quantified with the
chromospheric $S$-index \citep{wright_chromospheric_2004}.  We only
examined the HIRES velocities for this step, since they are the main
source of signal for our analysis.  First, we subtracted the orbital
solution from the Keck-HIRES velocities.  Then, following
\citet{bryan_statistics_2016,bryan_excess_2019}, we calculated the
Spearman rank correlation coefficient between the $S$-index and the
orbit-subtracted velocities.  We found a correlation coefficient of
0.16. This correlation is not statistically significant; the
corresponding $p$-value is 0.65.  Furthermore, inspection of the
$S$-index timeseries showed no secular or sinusoidal trends, as would
be expected if we were observing a long-term magnetic activity cycle.
The $S$-index values are included in {\bf Table X}.  We conclude that
it is highly unlikely that the linear trend is caused by
stellar activity.

@@

Given a linear radial velocity trend, we can place lower-limits on the
mass and semi-major axis of additional bodies in the system.  As a
quick estimate of the minimum mass required to explain the linear
trend in WASP-4, we turn to \citet{feng_california_2015}.  As they
discuss, the scenario that yields the minimum companion mass for a
system with a linear trend is a companion with $e\approx0.5$ and
$\omega=90^\circ$.  From their Equation~1,
\begin{equation}
 M_{\rm min} \approx 0.0164\,M_{\rm Jup}
  \left( \frac{\tau}{{\rm yr}} \right)^{4/3}
  \left| \frac{\dot{\gamma}}{{\rm m\,s^{-1}\,yr^{-1}}} \right|
  \left( \frac{M_\star}{{M_\odot}} \right)^{2/3},
\end{equation}
where $\tau$ is the observing baseline.
For WASP-4, this yields $M_{\rm min} = 4.7 M_{\rm Jup}$.  Higher mass
companions are allowed, presuming that they orbit further from the
star; at fixed $\dot{\gamma}$, $M_{\rm comp} \propto a^2$
\citep{torres_substellar_1999,liu_crossing_2002}.

@@

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is clearly false, we do not readily have
access to the planetary and stellar atmosphere models needed for the
consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We therefore adopted the effective temperatures and
bolometric luminosities from the \citet{baraffe_evolutionary_2003} and
MIST isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting limits are shown in
Figure~\ref{fig:mass_sma}.

@@

We assigned each companion a mass and semimajor axis from log-uniform
distributions within the grid cell. We drew the inclination from a
uniform distribution in $\cos i$.  For companion masses less than
$10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity distribution from the power-law $p_e
\propto e^\eta$ reported by \citet{moe_mind_2017} in their
Equation~17.  We emphasize that the long-period exoplanet and
long-period binary eccentricity distributions are quite different: the
exoplanet distribution is ``bottom-heavy'', with eccentricities
preferentially close to zero. The binary star distribution is
``top-heavy'', with eccentricities closer to one.  We chose a mass
cutoff of $10\,M_{\rm Jup}$ to separate the two regimes based on the
bound observed by \citet{schlaufman_evidence_2018} between giant
planets, presumably formed through core accretion, and brown dwarfs,
presumably formed via gravitational instability.

For each simulated companion, we then drew a sample from the converged
chains of our initial model of WASP-4b, plus its linear trend. We
subtracted the planet's orbital solution, leaving RV data points with
a linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the fixed instrument offsets and
jitters from the MCMC chains, we then performed a maximum likelihood
fit for the time and argument of periastron of the outer simulated
companion.  We converted the resulting $128\times128\times512$ cube of
best-fit log-likelihood values to probabilities, and averaged over the
samples in each grid cell to derive a probability distribution in mass
and semi-major axis.  Figure~\ref{fig:mass_sma} shows the result in
grayscale.

@@

The corresponding requirements for the companion causing the
acceleration are that it is likely either a brown-dwarf or low mass
star, and that it is between $10$-$100\,{\rm AU}$ from the host star
(Figure~\ref{fig:mass_sma}).  Given such a mass, this companion could
at one time have influenced the orbital evolution of the inner giant.
The fact that most hot Jupiters have similar massive outer
companions \citep{knutson_friends_2014,bryan_statistics_2016} is
circumstantial evidence for certain high-eccentricity formation
channels (see \citealt{dawson_johnson_2018}).
Further radial velocity monitoring should eventually reveal the
orbital parameters and minimum mass of WASP-4's companion.

@@

We identified WASP-4b's decreasing orbital period as part of a search
for tidal orbital decay.  However, most hot Jupiters have companions
outside of $1\,{\rm AU}$ with super-Jovian masses
\citep{knutson_friends_2014}.  Line-of-sight accelerations are
correspondingly relatively common in hot Jupiter systems.

To evaluate the importance of these effects in future transit timing
searches, we collected the linear radial velocity trends reported by
\citet{knutson_friends_2014}, and computed the expected orbital period
derivatives $\dot{P}_{\rm RV} = \dot{v}_{\rm r} P/c$ for all the
systems.  The results are given in {\bf Table~X}, and visualized for
hot Jupiters with significant linear trends in
Figure~\ref{fig:pdot_pop}.

Including WASP-4b, 16 of 51 hot Jupiters surveyed by
\citet{knutson_friends_2014} show a non-zero radial velocity trend.
About half are positive, and half are negative.  Therefore around 1 in
3 hot Jupiters are expected to show period changes commensurate with
WASP-4 due to acceleration by outer companions.  Half of these will be
period decreases, and will be an astrophysical false positives in
searches for tidal orbital decay.

For hot Jupiters that have been monitored for over 10 years, secular
changes in the orbital period be measured to a precision of $\lesssim
10~{\rm ms}\,{\rm yr}^{-1}$ ({\it e.g.}, K.~Patra et al{.} 2020,
submitted).

%FIXME
As observations from TESS continue...
As the Transiting Exoplanet

Continued long-term radial
  velocity monitoring of hot Jupiters is therefore essential to
  distinguish tidal orbital decay from line-of-sight accelerations.

@@

>>>>

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently indicated that its orbital period is
  decreasing.  The period change could be caused by tidal orbital
  decay, orbital precession, or light-travel time effects.  We present
  new radial velocity measurements of WASP-4, acquired with
  Keck-HIRES.  The data show that the system is accelerating towards
  the Earth at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
  m}\,{\rm s}^{-1}\,{\rm yr}^{-1}$.  The implied period decrease
  probably explains WASP-4's changing orbital period as a light-travel
  time effect.  Combining the radial velocities with new limits from
  speckle imaging, we find that the system's acceleration is likely
  caused by a 10-200$\,M_{\rm Jup}$ companion with semi-major axis
  between 10-100$\,$AU.  Using statistics from
  \citet{knutson_friends_2014}, about 1 in 6 hot Jupiters are expected
  to show similar period decreases as WASP-4, due to acceleration by
  outer companions.  We find that these period decreases will be
  increasingly measureable in coming years, because the precision of
  period derivative measurements scales as the fourth power of the
  observing baseline.  Continued radial velocity monitoring
  of hot Jupiters is therefore essential to distinguish tidal orbital
  decay from line-of-sight accelerations.
\end{abstract}

@@

The orbits of most hot Jupiters are unstable to tidal decay
\citep{counselman_outcomes_1973,hut_stability_1980,rasio_tidal_1996,levrard_falling_2009,matsumura_tidal_2010}.
The relevant issue is whether the timescale for tidal orbital decay is
shorter or longer than the timescale for main-sequence stellar
evolution.  This question depends on the uncertain rate at which
friction inside the star can damp the energy of tidal oscillations and
thereby shrink the orbit (as reviewed by \citealt{Mazeh2008} and
\citealt{ogilvie_tidal_2014}).

Indirect studies of age and angular momentum indicators including the
hot Jupiter semi-major axis distribution, host star spin rates, and
galactic velocity dispersions have led to estimates for the
inspiral timescale that vary from much less to much greater than
the main-sequence evolution time  ({\it e.g.},
\citealt{jackson_observational_2009}, \citealt{teitler_why_2014},
\citealt{penev_empirical_2018}, \citealt{cameron_hierarchical_2018},
\citealt{hamer_schlaufman_2019}).  Direct
measurements of tidal orbital decay through transit and occultation
timing could provide an empirical solution.  For instance, combined
transit timing and radial velocity measurements for WASP-12b have
shown that its secular period decrease of $\approx$30 milliseconds per
year is almost certainly due to tidal orbital decay
\citep{maciejewski_departure_2016,patra_2017,yee_orbit_2020}.

This study highlights a point that, though obvious, has perhaps not
yet received due attention.  The point is that observational programs
aimed at identifying orbital decay in hot Jupiters through transit
timing will be crippled without concurrent long-term radial velocity
monitoring.  The reason is that line-of-sight accelerations due to
outer companions \citep[{\it e.g.},][]{agol_detecting_2005} and tidal
orbital decay both initially manifest identically in transit times, as
a non-zero period derivative.  Massive outer companions to hot
Jupiters are the norm; \citet{bryan_statistics_2016} calculated an
occurrence rate of $60.9^{+5.2}_{-5.6}\%$ for outer companions to hot
Jupiters with masses from 1-20$M_{\rm Jup}$ and semi-major axes from
5-100$\,$AU.  Therefore it could very well be that more hot Jupiters
will show shrinking orbital periods due to long term accelerations
than due to tidal orbital decay.

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that appears to be decreasing by about 10 milliseconds
per year.  We discovered the timing variations using data from TESS
\citep{ricker_transiting_2015}
and a decade of ground-based observations \citep[][hereafter
\citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.  Thereafter,
\citet{southworth_transit_2019} reported 22 new transit times for the
system, and found an updated decay rate of $\dot{P} = -9.2 \pm 1.1$
milliseconds per year. The \citeauthor{southworth_transit_2019} decay
rate was $\approx$3$\sigma$ less rapid than that found by
\citetalias{bouma_wasp4b_2019}, but the conclusions of the studies
were otherwise similar.  A separate study by \citet{baluev_2019}
reported additional archival transit light curves of WASP-4b, most
notably from TRAPPIST, and also from select other
observers.  \citet{baluev_2019} pointed out that when using
lower-precision subsets of the available transit data, the case for a
decreasing period worsened.  We agree with this statement.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the five available
HIRES radial velocities suggested a weak ($\approx$$2\sigma$) linear trend
\citep{knutson_friends_2014}.  Our new measurements reveal a
line-of-sight acceleration of $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m\,s^{-1}\,yr^{-1}}$.  This
translates to an expected period decrease from the light-travel time
effect of -5.9 milliseconds per year---about commensurate with what is
observed from transits.  In the following, we avoid the term
``R{\o}mer delay'', which typically signifies arrival time delay due
to observatory motion. We are discussing the Doppler effect seen by a
stationary observer for a source with constant line-of-sight
acceleration.

In the following, \S~\ref{sec:observations} collects the available
transit data and presents the new radial velocity and speckle imaging
observations.  \S~\ref{sec:analysis} analyzes the data, and finds that
they yield a picture in which the WASP-4 system is accelerating
towards our line-of-sight, likely due to the pull of a brown or
M-dwarf companion.  \S~\ref{sec:discussion} places this result in the
broader context of orbital decay searches, and points out that
line-of-sight accelerations, {\it i.e.}, ``false positive orbital
decay signals'', are relatively common in the hot Jupiter population.
\S~\ref{sec:conclusions} offers concluding remarks.

@@

Table~1 lists the transit times we collected for our analysis.  We
include data from the peer-reviewed literature for which {\it (i)} the
analysis was based on observations of a single transit, {\it (ii)} the
midpoint was fitted as a free parameter, and {\it (iii)} the time
system specified both the leap second correction (TDB or UTC) and also
whether any barycentric or heliocentric corrections had been
performed.

@@

Additional timing measurements were also recently made available by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations taken professional and amateur observers.
We included twelve of their ``high quality'' transit times from
TRAPPIST (six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had or had not been performed.

@@

The complete set of radial velocity observations is given in 
Table~2.  Along with the 2010-2019 HIRES observations, there are also
earlier measurements from CORALIE and HARPS.  Following
\citetalias{bouma_wasp4b_2019}, we included the CORALIE measurements
from \citet{wilson_wasp-4b_2008} and \citet{triaud_spin-orbit_2010},
using the homogeneous radial velocities calculated by the latter
authors. We included the HARPS values reported by
\citet{pont_determining_2011} and \citet{husnoo_observational_2012}.
We omitted the HARPS data points taken over three nights by
\citet{triaud_spin-orbit_2010} for Rossiter-McLaughlin observations
because they have a systematic offset from the remaining datasets.

@@

We considered two models for the observed transit times.  The first
model assumes a constant orbital period $P$ on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes that the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
  \label{eq:quadratic_model}
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the observed times.  This helps
to reduce the covariance between $t_0$ and $P$.  A third possible
model that we did not consider for reasons that will become apparent
is a precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fit constant period model.  The best-fitting constant-period
model has 91 degrees of freedom, $\chi^2 = 276$,  and $\chi^2_{\rm
red}=3.0$.  The best-fitting quadratic model has 90 degrees of freedom,
$\chi^2 = 183$, and $\chi^2_{\rm red}=2.0$.  The difference in the
Bayesian information criteria (BIC) between the linear and quadratic
and models is $\Delta {\rm BIC} = 89.1$, strongly favoring the
quadratic case \citep{kass_bayes_1995}.

@@

One approach to analyzing such data would be to introduce a fudge
factor to inflate the transit measurement uncertainties, and lower the
reduced $\chi^2$.  We do not think that such an approach is warranted,
since it would not change the result that the quadratic model is
strongly preferred.

The best-fit period derivative in the quadratic model is
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$3$\sigma$ larger than the rate
of period decrease we reported earlier \citep[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of
the new data from \citeauthor{southworth_transit_2019}
and \citeauthor{baluev_2019}  The other best-fit
transit timing model parameters are reported in Table~3.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity residuals of WASP-4.} The best-fit
    Keplerian orbit of WASP-4b has been subtracted.  The linear trend
    inferred from the RV data is shown with a black line, with
    $1\sigma$ errors in gray.  The trend that would be needed to
    produce the period decrease seen in transits ($\dot{P}= - 8.64 \pm
    0.89~{\rm ms}\,{\rm yr}^{-1}$) is indicated with the purple dotted
    line.  The four new RV measurements presented in this work
    increased the significance of the radial velocity trend from
    $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period derivative expected from
radial velocities is
\begin{align}
  \dot{P}_{\rm\,RV}
  = \frac{\dot{v}_{\rm r} P}{c}
  = -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.

An important consideration is whether the measured RV trend is
correlated with stellar activity.  We investigate this by analyzing
WASP-4's emission in the Ca II H \& K lines, as quantified with the
chromospheric $S$-index \citep{wright_chromospheric_2004}.  We only
examined the HIRES velocities for this step, since they are the main
source of signal for our analysis.  First, we subtracted the orbital
solution from the Keck-HIRES velocities.  Then, following
\citet{bryan_statistics_2016,bryan_excess_2019}, we calculated the
Spearman rank correlation coefficient between the $S$-index and the
orbit-subtracted velocities.  We found a correlation coefficient of
0.16. This correlation is not statistically significant; the
corresponding $p$-value is 0.65.  Furthermore, inspection of the
$S$-index timeseries showed no secular or sinusoidal trends, as would
be expected if we were observing a long-term magnetic activity cycle.
The $S$-index values are included in Table~2.  We conclude that
it is highly unlikely that the linear trend is caused by
stellar activity.

@@

Given a linear radial velocity trend, we can place lower-limits on the
mass and semi-major axis of additional bodies in the system.  As a
quick estimate of the minimum mass required to explain the linear
trend in WASP-4, we turn to \citet{feng_california_2015}.  As they
discuss, the scenario that yields the minimum companion mass for a
system with a linear trend is a companion with $e\approx0.5$ and
$\omega=90^\circ$.  From their Equation~1,
\begin{equation}
 M_{\rm min} \approx 0.0164\,M_{\rm Jup}
  \left( \frac{\tau}{{\rm yr}} \right)^{4/3}
  \left| \frac{\dot{\gamma}}{{\rm m\,s^{-1}\,yr^{-1}}} \right|
  \left( \frac{M_\star}{{M_\odot}} \right)^{2/3},
\end{equation}
where $\tau$ is the observing baseline.
For WASP-4, this yields $M_{\rm min} = 4.7 M_{\rm Jup}$.  Higher masses
are allowed for companions that orbit further from the
star: at fixed $\dot{\gamma}$, $M_{\rm comp} \propto a^2$
\citep{torres_substellar_1999,liu_crossing_2002}.

@@

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is a bold simplification, we do not readily have
access to the planetary and stellar atmosphere models needed for the
consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We therefore adopted the effective temperatures and
bolometric luminosities from the \citet{baraffe_evolutionary_2003} and
MIST isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting limits are shown in
Figure~\ref{fig:mass_sma}.

@@

We assigned each companion a mass and semimajor axis from log-uniform
distributions within the grid cell. We drew the inclination from a
uniform distribution in $\cos i$.  For companion masses less than
$10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity distribution from the power-law $p_e
\propto e^\eta$ reported by \citet{moe_mind_2017} in their
Equation~17.  We emphasize that the long-period exoplanet and
long-period binary eccentricity distributions are quite different: the
exoplanet distribution is ``bottom-heavy'', with eccentricities
preferentially close to zero. The binary star distribution is
``top-heavy'', with eccentricities closer to one.  We chose a mass
cutoff of $10\,M_{\rm Jup}$ to separate the two regimes based on the
bound observed by \citet{schlaufman_evidence_2018} between
giant planets and brown dwarfs, though this value is also close to the
$13\,M{\rm Jup}$ required for a solar-metallicity body to
significantly burn deuterium \citep{burrows_nongray_1997}.

For each simulated companion, we then drew a sample from the converged
chains of our initial model of WASP-4b. We
subtracted the planet's orbital solution, leaving RV points with
a linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the fixed instrument offsets and
jitters from the MCMC chains, we then performed a maximum likelihood
fit for the time and argument of periastron of the outer simulated
companion.  We converted the resulting $128\times128\times512$ cube of
log-likelihood values to probabilities, and averaged over the
samples in each grid cell to derive a probability distribution in mass
and semi-major axis.  Figure~\ref{fig:mass_sma} shows the result.

@@

The corresponding requirements for the companion causing the
acceleration are that it is likely either a brown-dwarf or low mass
star, orbiting between $10$-$100\,{\rm AU}$ from the host star
(Figure~\ref{fig:mass_sma}).  Given such a mass, this companion could
at one time have influenced the orbital evolution of the inner giant.
The fact that most hot Jupiters have similar massive outer
companions \citep{knutson_friends_2014,bryan_statistics_2016} is
circumstantial evidence for certain high-eccentricity formation
channels (see \citealt{dawson_johnson_2018}).
Further radial velocity monitoring should eventually reveal the
orbital parameters and minimum mass of WASP-4's companion.

@@

@@

We identified WASP-4b's decreasing orbital period as part of a search
for tidal orbital decay.  However, most hot Jupiters have companions
outside of $5\,{\rm AU}$ with super-Jovian masses
\citep{knutson_friends_2014,bryan_statistics_2016}.  Line-of-sight accelerations are
correspondingly common in hot Jupiter systems.

To evaluate the importance of these effects for future transit timing
analyses, we collected the linear radial velocity trends reported by
\citet{knutson_friends_2014}, and computed the expected orbital period
derivatives $\dot{P}_{\rm\,RV} = \dot{v}_{\rm r} P/c$ for each
system.  The results are given in Table~4, and visualized for
hot Jupiters with significant ($>$$3\sigma$) linear trends in
Figure~\ref{fig:pdot_pop}.

Including WASP-4b, 16 of 51 hot Jupiters surveyed by
\citet{knutson_friends_2014} show a non-zero radial velocity trend.
Therefore around 1 in
3 hot Jupiters is expected to show period changes commensurate with
WASP-4 due to acceleration by outer companions.  Half of these will be
period decreases, and will be an astrophysical false positive in
searches for tidal orbital decay.

\subsection{At what rate is the measurement precision of d$P$/d$t$
increasing?}
\label{sec:fisher}
